---
title: "Introduction to structReport"
author: 
- name: Gavin R Lloyd
  affiliation: Phenome Centre Birmingham, University of Birmingham, UK
  email: g.r.lloyd@bham.ac.uk
- name: Ralf J Weber
  affiliation: Phenome Centre Birmingham, University of Birmingham, UK
  email: r.j.weber@bham.ac.uk
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2  
    number_sections: true  
    toc_float: true
package: structReport
vignette: >
  %\VignetteIndexEntry{Introduction to structReport}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---
  
```{r, include=FALSE}
  knitr::opts_chunk$set(
  dpi=72,
  fig.width=5,
  fig.height=5.5
  )
  set.seed(57475)
  library(structToolbox)
  library(structReport)
```
  
# Introduction
The aim of this vignette is to use demonstrate the use of `structReport` to generate
simple markdown documents for data analysis workflows implemented using the `struct` framework.
  
# Getting started
The latest version of `structReport` compatible with your current R version can be installed using `BiocManager`.
  
```{r, eval=FALSE, include=TRUE}
# install BiocManager if not present
if (!requireNamespace("BiocManager", quietly = TRUE))
install.packages("BiocManager")

# install structReport and dependencies
BiocManager::install("structReport",dependencies=TRUE)
  
```
  
# Report building blocks
In `structReport` we have defined a simple report structure that forms the basis 
of the package. 
  
## Document settings
A key component of markdown documents is the yaml header, where various settings 
are used to define e.g. title, author, document format etc. In `structReport`
this has implemented as a `struct_report` object. The `struct_report` object includes all
of the standard `struct` slots as well as slots used to define the yaml content of 
a report. Every report in `structReports` will therefore begin with a `struct_report` object.
In addition to the standard struct slots and the yaml slots there is a one additional slot:
  
- `sections`: A list of `report_section` objects defining the content of the report.

`report_section` objects are the second building block of a report and are discussed in the following section.

```{r}
# The first building block of a report
R = struct_report(
  title='Example report',
  author = 'Mr Happy',
  format = 'html_document',
  toc = TRUE,
  toc_depth = 2,
  sections= list()
)
```
  
## Report sections
  
The `report_section` object is used to define the content of sections to appear in a report. `report_section` objects
have three slots in addition to the standard ones defined by the `struct` package:
  
- `object`: The `struct` object this section refers to.
- `markdown`: Path to a markdown child document that will be knit into the main document.
- `subsection`: A list of `report_section` objects that are subsections of the current section.
  
  
```{r}
RS = report_section(
  object=DatasetExperiment(), # or model, iterator, chart etc
  markdown='default_object.Rmd',
  subsection = list()
)
```

Report sections can be added to a report creating a list of sections and assigning them to
the `sections` slot of `struct_report` object for the report.

```{r}
R = struct_report(
  title='Example report',
  author = 'Mr Happy',
  format = 'html_document',
  toc = TRUE,
  toc_depth = 2,
  sections=list(
    RS = report_section(
      object=DatasetExperiment(), # or model, iterator, chart etc
      markdown='default_object.Rmd',
      subsection = list()
    )
  )
)
```

Alternatively the "+" operator can be used to add sections to a report.

```{r}
R = struct_report(
  title='Example report',
  author = 'Mr Happy',
  format = 'html_document',
  toc = TRUE,
  toc_depth = 2) +
  report_section(
    name = 'Section 1',
    object=DatasetExperiment(), 
    markdown='default_object.Rmd'
  ) +
  report_section(
    name = 'Section 2',
    object=DatasetExperiment(), 
    markdown='default_object.Rmd'
  ) 
```

Subsections can be included by assigning `report_section` objects to the `subsection` slot of
other `report_section` objects.

```{r}
R = struct_report(
  title='Example report',
  author = 'Mr Happy'
) +
  report_section(
    name = 'Section 1',
    subsection = list(
      report_section(
        name = 'Section 1.1')
    )
  ) +
  report_section(
    name = 'Section 2'
  ) 
```

Internally `structReport` uses two markdown documents to automatically handle the 
setting of header levels and insertion of sections. 

## Markdown templates
The file specified in the `markdown` slot of a `report_section` will be inserted as 
a child document into the report as the output for the section. A basic "default_object.Rmd" is provided
but can be replaced by a custom markdown document if you wish. `strucReports` makes the 
object assigned to the `object` slot of a `report_section` available for use 
within the markdown, from which you can extract names, descriptions, parameter values 
outputs etc.

The default markdown template is very basic and doesnt require any input parameters.
However, because `struct_section` is derived from `struct_class` you can extend 
it and include input parameters to add custom options to your report e.g. font.

# Building the report
When the `struct_report` is ready and contains all the sections you need the
`build_report` method can be used to generate the report.

## Creating reports from struct objects
`structReport` defines a `as_report_section` method to enable you to quickly
generate a basic report structure from `struct` based objects. 

```{r}
# a struct model object for conducting PCA
M = PCA()

# generate a report section from the model
RS = as_report_section(M)
RS
```

The generated `report_section` object has already been populated with default values from
the model object.

**NOTE: The report creates a snapshot of the model at the moment the report object is created.
The model should therefore be trained _before_ creating the model object, or the results will not
be available to the report.**

## Default Rmd template
The default Rmd template is very basic:

- The name slot of the report section is used as the section heading.
- The description slot of the report section is used to generate the first paragraph.
- If the object is a chart object, then the object from the _parent_ section is used to plot the chart.
This means that charts can only be subsections of a report section that has a model object that
is appropriate for the chart e.g. a PCA scores plot needs a PCA model object in the parent section.

# Example
We demonstrate this for the Iris dataset and use `structToolbox` to implement a 
simple workflow to conduct a PCA.

First, we load the data, create our model and some charts objects.

```{r}
# Dataset
D = iris_DatasetExperiment()

# PCA model
M = mean_centre() + PCA(number_components = 4)
# apply the model or the results wont be reflected in the report.
M = model_apply(M,D) 

# scree plot
C1 = pca_scree_plot()

# scores plot
C2 = pca_scores_plot(factor_name = 'species')

```

Now we create our report structure using the provided objects.

```{r}
## start by setting yaml
R = struct_report(
  title='Example report',
  author= 'Mr Happy'
) 

## create a section describing the data by using the struct object name and description
S1 = as_report_section(D)

# create a section for our PCA model sequence
# add some details to the model sequence that will automatically included when 
# the report section is generated.
M$name = 'PCA workflow'
M$description = paste0('A model sequence consiting of two steps: 1) mean ',
  'centring and 2) PCA. Eaxch step is described in the following subsections.')

# include subsections for each chart
S2 = as_report_section(M,
  subsection = list(
    as_report_section(C1),
    as_report_section(C2)
  )
)

# combine into a report
R = R + S1 + S2
R
```

The final object is a report with two sections, and the second section has two
subsections, one for each chart.

An alternative, and maybe more transparent, approach to create the same report is
to create each section explicitly, instead of the using the `as_report_section` method.
This makes it easier to add custom headers and descriptions to the report sections but 
requires the model sequence to be split manually e.g. by indexing to generate a 
section for each step.

```{r}
R = struct_report(
  title='Example report',
  author= 'Mr Happy'
) +
  report_section(
    name = D$name,
    description = D$description,
    object = D
  ) +
  report_section(
    name = 'PCA workflow',
    description = paste0('A model sequence consiting of two steps: 1) mean ',
  'centring and 2) PCA. Eaxch step is described in the following subsections.'),
    subsection = list(
      report_section( # mean centre section
        object=M[1],
        name = M[1]$name,
        description = M[1]$description
      ),
      report_section( # pca section
        object=M[2],
        name = M[2]$name,
        description = M[2]$description,
        subsection = list( 
          report_section(
            object=C1, # pca scree plot
            name = C1$name,
            description = C1$description
          ),
          report_section(
            object=C2, # pca scores plot
            name = C2$name,
            description = C2$description
          ) 
        ) 
      ) 
    )
  )

R

```

Calling the `build_report` method on `R` will use `rmarkdown::render` to generate
a document of the desired format from the markdown templates set for each section.

```{r, eval=FALSE}

build_report(R,'example.docx')

```


# Cow/sheep example
This data is a more complex example involving numerous steps.

## Dataset
The dataset for this example is `MTBLS79` from the `struct`. It is a  Direct 
infusion mass spectrometry metabolomics dataset available from Metabolights
(https://www.ebi.ac.uk/metabolights/MTBLS79). The data has already been 
signal/batch corrected. We will use the filtered data and apply some preprocessing
before exploratory analysis. 

```{r,warning=FALSE,message=FALSE}
DE = MTBLS79_DatasetExperiment(filtered = TRUE)
DE
```

Now we create a single model sequence of all processing steps. These steps are described 
in detail in the vignette for `structToolbox`. When we generate the report later 
we will incorporate text from that vignette into the report.

```{r,warning=FALSE,message=FALSE}

M = pqn_norm(
      qc_label='QC',
      factor_name='sample_type'
    ) + 
    knn_impute(
      neighbours=5
    ) +
    glog_transform(
      qc_label='QC',
      factor_name='sample_type'
    ) +
    mean_centre(
    ) +
    PCA(
      number_components = 2
    )

# apply the sequence to the data
M = model_apply(M,DE)
```

Now we create a number of chart that we'll include in the final report.

```{r}

# Dataset
CS1a = mv_boxplot(factor_name='class',by_sample = TRUE,label_outliers = FALSE)
CS1b = mv_boxplot(factor_name='class', by_sample = FALSE,label_outliers = FALSE)

# pmp
CS3a = pqn_norm_hist()

# PCA by class
CS4a = pca_scores_plot(
      factor_name=c('sample_rep','class'),
      ellipse='none'
)

# PCA by batch
CS4b = pca_scores_plot(
      factor_name=c('batch'),
      ellipse='none'
)

```

Now we have everything we need to build our report. We'll add sections for relevant
steps in the model sequence, and provide descriptions for each. Charts will
added as subsections where appropriate.

We will break down each part of the report before putting it all together at the end.
First we define the report yaml using a `struct_report` object.
```{r}
R = 
  struct_report(
    title = 'A typical workflow for processing and analysing mass spectrometry-based metabolomics data.',
    author = 'G.R. Lloyd',
    format = "word_document"
  ) 
```

First a simple section with no model defined. The default Rmd template uses
the `name` slot as the header and the `description` as the contents.

```{r}
S1 = report_section(
      name = 'Introduction',
      description = paste0(
      'This report provides an overview of a structToolbox workflow implemented ',
      'to process (e.g. filter features, signal drift and batch correction, ',
      'normalise and missing value imputation) mass spectrometry data. The ',
      'workflow exists of methods that are part of the Peak Matrix Processing ',
      '(pmp) package, including a range of additional filters that are ',
      'described in Kirwan et al., 2013, 2014.'
      )
  ) 
```

Next we create a section to describe the data, and produce some summary plots. 
The plots are included as subsections of the main section. The charts use the 
object from their parent section as input for the plot (`DE` in this case).

For the chart subsections we set `name` = character(0) which using the default
Rmd template means no header will be generated for these sections i.e. the
charts will appear under the section header of the parent section.

```{r}
S2 =  report_section(
        name = 'Data preparation',
        object = DE,
        description = paste0(
      DE$description,
      "\n\n",
      "The `MTBLS79_DatasetExperiment` object included in the `structToolbox` ",
      "package is a processed version of the MTBLS79 dataset available in peak ",
      "matrix processing (`r Biocpkg('pmp')``) package.",
      "\n\n",
      "Here, we load in the data from the `struct` package 'unfiltered' and apply ,",
      "the filtering in a model sequence to demonstrate the use of `structReport` ",
      "with more complex analysis workflows."
    ),
      subsection=c(
      report_section(
        name=character(0),
        description='A boxplot of missing values per sample for each group.',
        object=CS1a
      ),
      report_section(
        name=character(0),
        description='A boxplot of missing values per feature for each group.',
        object=CS1b
      )
    )
  ) 
```

In this section we describe the peak matrix processing steps applied and include 
some charts for each step. We use the object `name` and `description` in the report
instead of writing text ourselves. We have also incorporated bulleted list formatting
into the description which will be interpreted when the report is built.
```{r}

S3 = report_section(
        name = 'Normalisation, imputation and scaling',
        description = paste0(
          'We will apply a number of common pre-processing steps to the filtered ',
          'peak matrix that are identical to steps applied in are described in ',
          'Kirwan et al. 2013, 2014.',
          '\n\n',
          '- Probabilistic Quotient Normalisation (PQN)\n',
          '- k-nearest neighbours imputation (k = 5)\n',
          '- Generalised log transform (glog)',
          '\n\n',
          'These steps prepare the data for multivariate analysis by accounting for ',
          'sample concentration differences, imputing missing values and scaling ',
          'the data.'
    ),
    subsection = c(
      report_section( # PQN
        name = M[1]$name,
        description = M[1]$description,
        object = M[1],
        subsection = 
          report_section(
            name=character(0),
            description=CS3a$description,
            object = CS3a
          )
      ),
      report_section( # imputation
        name = M[2]$name,
        description = M[2]$description,
        object = M[2]
      ),
      report_section( # glog
        name = M[3]$name,
        description = M[3]$description,
        object = M[3]
      )
    )
  )
```

In the final section we describe Exploratory analysis using PCA scores plots
to visualise the data by batch and sample group.

```{r}
S4  = report_section(
    name = 'Exploratory analysis',
    object = M[5],
    description = paste0(
      'Principal Component Analysis (PCA) can be used to visualise ',
      'high-dimensional data. It is an unsupervised method that maximises ',
      'variance in a reduced number of latent variables, or Principal Components.'
    ),
    subsection = list(
      report_section(
        name=character(0),
        object = C2,
        description = paste0(
        'This plot is very similar to Figure 3b of the original publication. ',
        'Sample replicates are represented by colours and samples groups ',
        '(C = cow and S = Sheep) by different shapes.'
        )
      ),
      report_section(
        name=character(0),
        object = C3,
        description = paste0(
        'Plotting the scores and colouring by Batch indicates that the ',
        'signal/batch correction was effective as all batches are overlapping.'
        )
      )
    )
  )
```
  
The final report can be converted to the chosen format (word docx in this case)
by using the `build_report` method.

```{r,eval = FALSE}
REPORT = R+S1+S2+S3+S4
build_report(REPORT,'MTBLS79_example.docx')
```










