---
title: "`r R$title`"
author: "`r R$author`"
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
```

# This is the basic template for a struct report document.
The report object is "R" in the workspace, and the current data is "D". There is
no model, as the idea is to iterate over the sections, with each section potentially
using a different model.

To function correctly with any report sections included in the object it should include
a section of r code that loops over and knits the report_sections.

The model should already have been trained.

```{r,echo=FALSE, include=TRUE,eval=length(R)>0}

# loop over child docs
out=NULL
counter=1
for (k in 1:length(R)) {
    # set the section object for current child
    S = R$sections[[k]]
    # set the model for this section
    MS=NULL
    if (!(is.null(R$sections[[k]]$model_name))) {
        MS = M[counter]
        counter=counter+1 # only increments if model_name isn't NULL
    }
  out <- c(out, knitr::knit_child(R$sections[[k]]$markdown,quiet=TRUE,envir=environment()))
}
```

```{r, echo=FALSE, results="asis",eval=length(R)>0}
cat(paste(out, collapse = '\n'))
```
